A Tutorial on Tailored Power Simulations for Experimental Study Designs with Generalized Linear Mixed Models

### 2023-09-13 docs/manuscript.Rmd
work on comments

<!-- Advantages of GLMMs over "simpler" models -->
As study designs become more complex<!--AK: I am not even sure whether this is true. Do study designs become more complex or do we become more aware of the pitfalls of underpowered studies? I would say it's the latter.-->, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data in HCI and other empirical disciplines because they offer a flexible framework for analyzing data with non-normal and categorical outcomes while accounting for both fixed and random effects [@kapteinUsingGeneralizedLinear2016; @fahrmeirRegressionModelsMethods2021].

@@

<!-- what are the existing options for a priori power analysis for GLMMs / where are existing tutorials limited -->
Currently available software packages for power analysis <!-- FP: because you removed the sentence on the difference between analytic and simulation based power analysis, it is now not clear why from now on we always talk about power simulation; TK: sorry if I have accidentally cut this distinction when merging versions, i think it is super important to have the definition between the two here.  --> do not include GLMMs, making it necessary to build data simulations tailored specifically to the study design [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022]. 
A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials only cover linear mixed models (LMMs) and focus on the most common designs (e.g., INSERT EXAMPLE <!--FP can you think of a good example?-->). This narrow focus provides little guidance for researchers faced with more complex study designs (e.g., INSERT EXAMPLE <!--FP can you think of a good example?-->) [@brownIntroductionLinearMixedEffects2021]. 
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality<!--AK: Normality of what?--> and the covariance structure among the random effects<!--AK: Unclear what this is-->. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying presumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions <!--AK: Again "assumptions".--> a practical challenge [@maxwellSampleSizePlanning2008].

>>>>

<!-- Advantages of GLMMs over "simpler" models -->
As study designs become more complex<!--AK: I am not even sure whether this is true. Do study designs become more complex or do we become more aware of the pitfalls of underpowered studies? I would say it's the latter. -->, researchers require more sophisticated statistical models to capture the nuanced relationships and hierarchical structures introduced by their study designs [@yarkoniGeneralizabilityCrisis2022]. GLMMs (also called multilevel models) are gaining increasing popularity in analyzing data in HCI and other empirical disciplines because they offer a flexible framework for analyzing data with non-normal and categorical outcomes while accounting for both fixed and random effects [@kapteinUsingGeneralizedLinear2016; @fahrmeirRegressionModelsMethods2021].

@@

<!-- what are the existing options for a priori power analysis for GLMMs / where are existing tutorials limited -->
Currently available software packages for power analysis <!-- FP: because you removed the sentence on the difference between analytic and simulation based power analysis, it is now not clear why from now on we always talk about power simulation; TK: sorry if I have accidentally cut this distinction when merging versions, i think it is super important to have the definition between the two here.  --> often do not include GLMMs [@westfallStatisticalPowerOptimal2014; @murayamaSummarystatisticsbasedPowerAnalysis2022], making it necessary to build data simulations tailored specifically to the study design. 
A number of tutorials have been published describing how to perform such simulation-based power analysis for multilevel models [@arendStatisticalPowerTwolevel2019; @lafitSelectionNumberParticipants2021; @zimmerSampleSizePlanning2022; @brysbaertPowerAnalysisEffect2018; @kumleEstimatingPowerGeneralized2021; @debruineUnderstandingMixedEffectsModels2021]. 
However, most of these tutorials focus on linear mixed models (LMMs) and the most common designs [but see @kumleEstimatingPowerGeneralized2021 for a tutorial that also covers more advanced settings]. This narrow focus provides limited guidance for researchers faced with more complex study designs especially when little prior knowledge about plausible effect sizes is available [see the discussion in @kumleEstimatingPowerGeneralized2021]. 
<!-- FP: es war nicht leicht das hier zu lösen: ich möchte ungern debruineUnderstandingMixedEffectsModels2021 als negativ beispiel hier nennen, das hat zwar nur lmms im hauptteil aber glmms im appendix und ist einfach ein super paper. außerdem hatte ich beim nochmal lesen den eindruck, dass wir kumleEstimatingPowerGeneralized2021 in unserem ganzen paper nicht genug credit geben. daher hab ich es jetzt so umgestellt dass wir nicht die negativbeispiele sondern das positivbeispiel aufzählen -->
The necessary assumptions for power simulations in GLMMs include assumptions about the distributional form of the outcome variable, the random effects, and the error structure. The distributional assumption specifies the family of distributions for the outcome variable, such as Gaussian, Poisson, or binomial. Assumptions about the random effects include the assumption of normality<!--AK: Normality of what?--> and the covariance structure among the random effects<!--AK: Unclear what this is-->. Additionally, assumptions about the error structure, such as independence or correlation, must be specified. Interpreting these assumptions entails understanding the underlying presumptions of the model and ensuring they align with the characteristics of the data being analyzed.
Existing tutorials often rely on heuristics for specifying variance components (e.g., the standard deviation of random intercepts) or assume that results from meta-analyses or data from pilot studies are available to determine plausible values for all model parameters. However, in practice, knowledge about those parameters from prior studies is often limited, which makes specifying assumptions <!--AK: Again "assumptions".--> a practical challenge [see the discussion in @maxwellSampleSizePlanning2008 and @kumleEstimatingPowerGeneralized2021].
